
```{r }
```

   GEOSTATISTICS   
--------------------
Clear the workspace

```{r }
rm(list=ls())
```

Load spatial packages

```{r }
library(sp)           # Data management
library(lattice)      # Data management
library(geoR)         # Geostatistics
library(gstat)        # Geostatistics
```

Functions for graphics 

```{r }
v.f <- function(x, ...){100-cov.spatial(x, ...)}
v.f.est<-function(x,C0, ...){C0-cov.spatial(x, ...)}
```

  EXPLORATORY ANALYSIS & VARIOGRAM ESTIMATION
Load meuse data set:
The meuse is a classical geostatistical data set used frequently
to demonstrate various geostatistical analysis steps.
The point data set consists of 155 samples of top soil heavy metal
concentrations (ppm), along with a number of soil and landscape variables.
The samples were collected in a flood plain of the river Meuse,
near the village Stein (The Netherlands).

```{r }
data(meuse)
```

Define the sample coordinates

```{r }
coordinates(meuse) <- c('x','y')
```

bubble plot(obj,zcol,...)
key.space=location of the key

```{r }
bubble(meuse,'zinc',do.log=TRUE,key.space='bottom')

dev.off()
```

river meuse

```{r }
data(meuse.riv)
meuse.lst <- list(Polygons(list(Polygon(meuse.riv)), "meuse.riv"))
meuse.sr <- SpatialPolygons(meuse.lst)
```

grid for prediction

```{r }
data(meuse.grid)
is(meuse.grid)
coordinates(meuse.grid) <- c('x','y')
meuse.grid <- as(meuse.grid, 'SpatialPixelsDataFrame')
```

plot all together

```{r }
image(meuse.grid, col = "lightgrey")
plot(meuse.sr, col = "grey", add = TRUE)
plot(meuse, add = TRUE)
title('meuse river geostatistical data')

dev.off()
```

Exploratory Analysis #
----------------------#
histogram of zinc variable

```{r }
hist(meuse$zinc, breaks=16, col="grey", main='Histogram of Zn', prob = TRUE, xlab = 'Zn')
```

highly skewed, transform to the log

```{r }
hist(log(meuse$zinc), breaks=16, col="grey", main='Histogram of log(Zn)', prob = TRUE, xlab = 'log(Zn)')
```

scatterplot of log(zinc) with respect to distance from the river 

```{r }
xyplot(log(zinc) ~ sqrt(dist), as.data.frame(meuse))
```

Negative correlation: lower distance from the river => higher level of zinc

```{r }
dev.off()
```

Estimating Spatial Correlation #
    Variogram Analysis         #
--------------------------------#
sample variogram (binned estimator)

```{r }
svgm <- variogram(log(zinc) ~ 1, meuse)
plot(svgm, main = 'Sample Variogram',pch=19)

dev.off()
```

N.B.  the notation "~ 1" stands for a single constant predictor
      (hp: spatially constant mean)
default decisions:
direction dependence, cutoff, lag width
the following

```{r }
plot(variogram(log(zinc) ~ 1, meuse),pch=19)
```

automatically decides to ignore direction: point pairs are merged on the
basis of distance to compute the empirical variogram

```{r }
plot(variogram(log(zinc) ~ 1, meuse, alpha = c(0, 45, 90, 135)),pch=19)
```

point pairs whose separation vector has a given direction are used in each
panel (not too many directions otherwise noise will increase)
Note: zonal anisotropy
cutoff distance: maximum distance up to which point pairs are considered
                 (default = bbox diagonal / 3)
lag width: width of distance intervals over which point pairs are averaged
           in bins (default = cutoff distance / 15)

```{r }
plot(variogram(log(zinc) ~ 1, meuse, cutoff = 1000, width = 1000/15),pch=19)
```

intervals can have different widths: to fix varying widths use the argument
boudaries

```{r }
plot(variogram(log(zinc) ~ 1, meuse, boundaries = c(0,200,seq(400,1500,100))),pch=19)
```

useful for data sets that have much information on short distance variability

```{r }
plot(variogram(log(zinc) ~ 1, meuse, cutoff = 1000, width = 1000/5),pch=19)
```

Variogram modeling
-------------------
list of parametric isotropic variogram models

```{r }
vgm()
```

in gstat, valid variogram models are constructed by using one or
combination of two or more basic variogram models
first argument of the function 'vgm' is partial sill,
then the desired model, then range, and finally nugget: 
vgm(sill, model, range, nugget)
some examples...

```{r }
vgm(1, "Sph", 300)
vgm(1, "Sph", 300, 0.5)
```

one can also add two or more models

```{r }
v1 <- vgm(1, "Sph", 300, 0.5)
v2 <- vgm(0.8, "Sph", 800, add.to = v1)
v2
```

this is only measurement error

```{r }
vgm(0.5, "Nug", 0)
```

weighted least squares fitting a variogram model to the sample variogram
STEPS:
1) choose a suitable model
2) choose suitable initial values for partial sill, range & nugget
3) fit the model using one of the possible fitting criteria

```{r }
v <- variogram(log(zinc) ~ 1, meuse)

plot(v,pch=19)
```

Linear behavior near the origin, growth not very fast 
Recall: both spherical and exponential model have a linear behavior near the
        origin but exponential model has a faster growth than the spherical one
=> we fit a spherical model
try reasonable initial values

```{r }
fit.variogram(v, vgm(1, "Sph", 800, 1))
```

try unreasonable initial values

```{r }
fit.variogram(v, vgm(1, "Sph", 10, 1))
```

due to high non linearity in the minimization problem,
starting from unreasonable initial values might cause fail to converge
plot of the final fit

```{r }
v <- variogram(log(zinc) ~ 1, meuse)
v.fit <- fit.variogram(v, vgm(1, "Sph", 800, 1))
plot(v, v.fit, pch = 19)
```

fitting method: non linear regression with minimization of weighted
sum of squares error. final value of the minimum

```{r }
attr(v.fit, 'SSErr')
```

how can we choose weights? argument fit.method in fit.variogram
fit.method = 1 : w = N_j
fit.method = 2 : w = N_j/gamma(h_j)^2
fit.method = 6 : w = 1
fit.method = 7 : w = N_j/h_j^2
one can also keep one of the parameters fixed, and fit only the others.
this is common for the nugget parameter, which may be hard to infer from data
when sample locations are regularly spread. Information may be derived from
measurement error characteristics for a specific device.
ex: fix the nugget variance to the value 0.06

```{r }
fit.variogram(v, vgm(1, "Sph", 800, 0.06), fit.sills = c(FALSE, TRUE))
```

the range parameters can be fixed using argument fit.ranges
maximum likelihood fitting of variogram models
- does not need the sample variogram
- can be performed through restricted maximum likelihood

```{r }
fit.variogram.reml(log(zinc)~1, meuse, model=vgm(0.6, "Sph", 800, 0.06))
v.fit
```

modeling anisotropy*

```{r }
v.dir <- variogram(log(zinc)~1,meuse,alpha=(0:3)*45) 
v.anis <- vgm(.6, "Sph", 1600, .05, anis=c(45, 0.3))

print(plot(v.dir, v.anis, pch=19))
```


                                #
         SPATIAL PREDICTION & KRIGING          #
                                #

Stationary Univariate Spatial Prediction (Ordinary Kriging)
-------------------------------------------------------------
Prediction at a single new location 

```{r }
s0.new=data.frame(x=179180, y=330100) # UTM coordinates
coordinates(s0.new)=c('x','y')
```

plot all together

```{r }
image(meuse.grid, col = "lightgrey")
plot(meuse.sr, col = "grey", add = TRUE)
plot(meuse, add = TRUE)
plot(s0.new, add = TRUE, col='red', lwd = 2)
title('meuse river geostatistical data')
```

Create a gstat object setting a spherical (residual) variogram
gstat(g.obj, id, formula, data, model, set,...)

```{r }
g.tr <- gstat(formula = log(zinc) ~ 1, data = meuse, model = v.fit)
```

ordinary kriging
Make the ordinary kriging prediction with the function: 
predict(obj, grid, BLUE=FALSE)
this gives the prediction of Y(s_0):

```{r }
predict(g.tr, s0.new)
```

variance > 0 (as expected)
Estimate the mean:use the argument 'BLUE'

```{r }
predict(g.tr, s0.new, BLUE = TRUE)
```

this gives the estimate of the mean
(trend component) under gls
consider a location where I OBSERVE data
this gives the prediction of Y(s_0)
in the first location (zero variance!)

```{r }
meuse[1,]
predict(g.tr, meuse[1,])
```

this gives the estimate of the mean
(drift component) under gls

```{r }
predict(g.tr, meuse[1,], BLUE = TRUE)
```

prediction over the entire grid

```{r }
lz.ok <- predict(g.tr, meuse.grid, BLUE = FALSE)

spplot(lz.ok)
```

Non-stationary Univariate Spatial Prediction (Universal Kriging)
-----------------------------------------------------------------
the hypothesis of spatially constant mean may be too restrictive!
we now use as covariate the square root of the distance from the river Meuse
to fit the variogram on the residuals, one should take into account 
the spatial dependence while estimating the trend component by using GLS
Create a gstat object setting a spherical (residual) variogram
gstat(g.obj, id, formula, data, model, set,...)

```{r }
meuse.gstat <- gstat(id = 'zinc', formula = log(zinc) ~ sqrt(dist),
                     data = meuse, nmax = 50, model=v.fit, set = list(gls=1))
meuse.gstat
```

Estimate the variogram from GLS residuals:

```{r }
?variogram.gstat
v.gls<-variogram(meuse.gstat)
plot(v.gls)

v.gls.fit <- fit.variogram(v.gls, vgm(1, "Sph", 800, 1))
plot(v.gls, v.gls.fit, pch = 19)
```

Update gstat object with variogram model

```{r }
meuse.gstat <- gstat(id = 'zinc', formula = log(zinc) ~ sqrt(dist),
                     data = meuse, nmax = 50, model=v.gls.fit, set = list(gls=1))
```

universal kriging:
I have to define the covariate in s_0

```{r }
s0.vec <- as.vector(slot(s0.new,'coords'))
```

distance to the river: calculate the distance between s0 and each point of
the river, then select the minimum

```{r }
s0.dist <- min(rowSums(scale(meuse.riv,s0.vec)^2)) 
s0.new <- as.data.frame(c(s0.new,s0.dist))
names(s0.new) <- c('x','y','dist')
coordinates(s0.new) <- c('x','y')
s0.new <- as(s0.new, 'SpatialPointsDataFrame')
s0.new
```

Function "predict" uses the residual variogram stored in the gstat
object to make the prediction

```{r }
predict(meuse.gstat, s0.new)
```

variance > 0 (as expected)
this gives the estimate of x(s_0)'*beta
(trend component) under gls

```{r }
predict(meuse.gstat, s0.new, BLUE = TRUE)
```

prediction over the entire grid

```{r }
lz.uk <- predict(meuse.gstat, meuse.grid, BLUE=FALSE)
```

estimate of the mean over the entire grid

```{r }
lz.uk.BLUE <- predict(meuse.gstat, meuse.grid, BLUE=TRUE)

spplot(lz.ok[,1], main = 'Ordinary Kriging, gstat')

spplot(lz.uk[,1], main = 'Universal Kriging, gstat')

spplot(lz.uk.BLUE[,1], main = 'Universal Kriging - drift , gstat')
```

Is the drift important to explain the variability of the response variable?
Let's compare the variogram of the data and of the residuals:

```{r }
plot(v$dist,v$gamma,xlab='distance',ylab='semivariance',pch=19,col='skyblue1',ylim=c(0,0.8))
curve(v.f.est(x, C0=v.fit[2,2]+v.fit[1,2], cov.pars=rbind(c(v.fit[2,2], v.fit[2,3]),c(v.fit[1,2], v.fit[1,3])), cov.model = c("spherical","pure.nugget")), from = 0.0001, to = 1600,
      xlab = "distance", ylab = expression(gamma(h)),
      main = "Variogram model",add=TRUE,col='skyblue1',lwd=2, ylim=c(0,110))
points(v.gls$dist,v.gls$gamma,xlab='distance',ylab='semivariance',pch=19,col='steelblue',ylim=c(0,0.8))
curve(v.f.est(x, C0=v.gls.fit[2,2]+v.gls.fit[1,2], 
              cov.pars=rbind(c(v.gls.fit[2,2], v.gls.fit[2,3]),c(v.gls.fit[1,2], v.gls.fit[1,3])), cov.model = c("spherical","pure.nugget")), from = 0.0001, to = 1600,
      xlab = "distance", ylab = expression(gamma(h)),
      main = "Variogram model",add=TRUE,col='steelblue',lwd=2, ylim=c(0,110))

graphics.off()
```


                                #
          EXERCISES FROM PAST EXAMS            #
                                #

One of the most relevant consequences of the eruption of volcan 
Eyjafjoll (in Iceland), in 2010, is the contamination by fluoride. 
The latter is due to the deposit on the ground of the ash released
in the atmosphere during the eruption.
The file "fluoruro.txt" reports the coordinates of 50 measurement sites
s_i, i=1,...,50, the corresponding concentrations of fluoride (ppm) F(s_i)
and the distance D.s_i of each site s_i from the crater of the volcano.
Denoting by delta a zero-mean, second-order stationary and isotropic random
field:
a) Estimate two empirical variograms, assuming the following models:
   F(s_i)=beta_0+delta(s_i) and 
   F(s_i)=beta_0+beta_1*D.s_i+delta(s_i). 
   Choose the most appropriate model for the observations.
b) Fit to the empirical variogram at point (a), a Gaussian model
   without nugget, via weighted least squares. Use as initial parameters: 
   sill=100, range=0.02. Report the estimates of sill and range.
c) Fit to the empirical variogram chosen at point (a), a spherical model
   without nugget, via weighted least squares. Report the estimates of sill 
   and range.
d) Compare the variograms estimated at points (b) and (c), with the empirical
   variogram at point (a). Given that the ash deposition is known to be
   a very regular phenomenon, which variogram model is the most appropriate?
e) Based on model (d), estimate the concentration of fluoride due to the eruption
   in the city of Raufarhofn (s0 = (0.3; 0.24), D.s0 = 0.1970) 
f) Based on model (d), estimate the concentration of fluoride at the same
   location, due to a possible new eruption of equivalent intensity, independent 
   of that of 2010.
 SOLUTION
_____________
Data import

```{r }
data=read.table(here::here('markdowns','lab_13_data','fluoruro.txt'))
names(data)[3]='f'
attach(data)
coordinates(data)=c('X','Y')
```

a) Estimate two empirical variograms, assuming the following models:
   F(s_i)=beta_0+delta(s_i) and 
   F(s_i)=beta_0+beta_1*D.s_i+delta(s_i). 
   Choose the most appropriate model for the observations.

```{r }
v=variogram(f ~ 1, data=data)
plot(v,pch=19)

v.t=variogram(f ~ D, data=data)
plot(v.t,pch=19)
```

b) Fit to the empirical variogram at point (a), a Gaussian model
   without nugget, via weighted least squares. Use as initial parameters: 
   sill=100, range=0.02. Report the estimates of sill and range.

```{r }
v.fit2 <- fit.variogram(v.t, vgm(100, "Gau", 0.02))
plot(v.t, v.fit2, pch = 3)
v.fit2
```

c) Fit to the empirical variogram chosen at point (a), a spherical model
   without nugget, via weighted least squares. Report the estimates of sill 
   and range.

```{r }
v.fit1 <- fit.variogram(v.t, vgm(100, "Sph", 0.08))
plot(v.t, v.fit1, pch = 3)
v.fit1
```

d) Compare the variograms estimated at points (b) and (c), with the empirical
   variogram at point (a). Given that the ash deposition is known to be
   a very regular phenomenon, which variogram model is the most appropriate?
We choose v.fit2
e) Based on model (d), estimate the concentration of fluoride due to the eruption
   in the city of Raufarhofn (s0 = (0.3; 0.24), D.s0 = 0.1970) 

```{r }
g.t <- gstat(formula = f ~ D, data = data, model = v.fit2)

D.s0=0.1970
s0=as.data.frame(matrix(c(0.3,0.24,D.s0),1,3))
names(s0)=c('X','Y','D')
coordinates(s0)=c('X','Y')

predict(g.t, s0, BLUE = FALSE)
```

f) Based on model (d), estimate the concentration of fluoride at the same
   location, due to a possible new eruption of equivalent intensity, independent 
   of that of 2010.

```{r }
predict(g.t, s0, BLUE = TRUE)

detach(data)
```

________________________________________________________________________________

The file radioville.txt reports the information on 158 control units
in the area around the nuclear power plant of Radioville.
At each site, available data consist of: radioactivity levels [Bq],
longitude [°N], latitude [°W] and type of soil [urban/vegetation].
Denoting by s_i the i-th site, by R the radioactivity level,
by eps a weakly stationary random field and by D a dummy 
urban/vegetation:
a) estimate the parameters of the linear model 
   R(s_i) = beta_0 + beta_1 D(s_i) + eps(s_i) 
   assuming for eps a spherical variogram without nugget, estimated
   via weighted least squares;
b) estimate the parameters of the linear model 
   R(s_i) = beta_0 + beta_1 D(s_i) + eps(s_i) 
   assuming for eps a spherical variogram with nugget, estimated
   via weighted least squares;
c) choose the best variogram model by comparing the fitted model 
   with the corresponding empirical variograms (report qualitative
   plots and the estimated variogram parameters)
d) on the basis of model (c), predict the radioactivity level at the
   parking lot of the shopping centre of Radioville (lon = 78.59, 
   lat = 35.34), and in the park of Radioville (lon = 77.6, 
   lat = 34.99);
e) estimate variance of prediction error at the same locations
   as at point d).
 SOLUTION
_____________

```{r }
data <- read.table(here::here('markdowns','lab_13_data','radioville.txt'),header=TRUE)
attach(data)
```

create dummy: 0 = urban, 1 = vegetation

```{r }
DUMMY <- rep(0,length(D))
DUMMY[which(D=='V')] <- 1
data <- data.frame(cbind(Bq,Long,Lat,DUMMY))
names(data) <- c('Bq','Long','Lat','D')
coordinates(data) <- c('Long','Lat')
```

point a)
fitting a variogram without nugget

```{r }
v <- variogram(Bq ~ D, data = data)
plot(v)
v.fit1 <- fit.variogram(v, vgm(1, "Sph", 0.5))
plot(v, v.fit1, pch = 3)
v.fit1
```

coefficient of the linear model: 
it sufficies to estimate the drift at two locations where we have observations,
with D=U and D=V
data[1,] = urbane
data[6,] = vegetation

```{r }
g.tr <- gstat(formula = Bq ~ D, data = data, model = v.fit1)
predict(g.tr, data[1,], BLUE = TRUE)
predict(g.tr, data[6,], BLUE = TRUE)
```

point b)
fitting a variogram with nugget

```{r }
v <- variogram(Bq ~ D, data = data)
plot(v)
v.fit2 <- fit.variogram(v, vgm(0.6, "Sph", 0.5, 0.1))
plot(v, v.fit2, pch = 3)
v.fit2
```

it sufficies to estimate the drift at two locations where we have observations,
with D=U and D=V
data[1,] = urbane
data[6,] = vegetation

```{r }
g.tr <- gstat(formula = Bq ~ D, data = data, model = v.fit2)
predict(g.tr, data[1,], BLUE = TRUE)
predict(g.tr, data[6,], BLUE = TRUE)
```

point d)
predict at 2 new locations: we use model 1 (without nugget)

```{r }
g.tr <- gstat(formula = Bq ~ D, data = data, model = v.fit1)
```

urbane : 78.59,35.34

```{r }
s0.new <- as.data.frame(matrix(c(78.59,35.34,0),1,3))
names(s0.new) <- c('lon','lat','D')
coordinates(s0.new) <- c('lon','lat')
predict(g.tr, s0.new)
```

vegetation : 77.69,34.99

```{r }
s0.new <- as.data.frame(matrix(c(77.69,34.99,1),1,3))
names(s0.new) <- c('lon','lat','D')
coordinates(s0.new) <- c('lon','lat')
predict(g.tr, s0.new)
```

